---
title: "Tutorials on Infectious Disease Models"
author: "Sen Pei"
date: "8/25/2020"
output:
  html_document:
    code_folding: hide
    df_print: paged
    toc: yes
    toc_float: yes
    theme: flatly
  pdf_document:
    toc: yes
  word_document:
    toc: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## SIR model
Variables: Total population $N$, susceptible population $S$, infected population $I$, recovered population $R$.

Parameters: The transmission rate $\beta$, average infectious period $D$ days.

Equations:

\begin{eqnarray}
  \frac{dS}{dt} &=& -\beta\frac{SI}{N},\\
  \frac{dI}{dt} &=& \beta\frac{SI}{N}-\frac{I}{D},\\
  \frac{dR}{dt} &=& \frac{I}{D}.
\end{eqnarray}

We use an Euler stepping scheme to discretize the equations and a Possion process to introduce stochasticity:

\begin{eqnarray}
  S(t+1) &=& S(t) - Pois(\beta\frac{S(t)I(t)}{N}),\\
  I(t+1) &=& I(t) + Pois(\beta\frac{S(t)I(t)}{N}) - Pois(\frac{I(t)}{D}),\\
  R(t+1) &=& R(t) + Pois(\frac{I(t)}{D}).\\
\end{eqnarray}


Function:

```{r SIR}
#SIR model
SIR<-function(N, S, I, beta, D, tmstep){
  #total popualtion: N; susceptible: S; infected: I; 
  #transmission rate: beta; average infectious period: D
  #simulation time: tmstep (days)
  
  #time series for susceptible, infected and newly infected population
  S_rec = I_rec = R_rec = newI_rec = matrix(0, tmstep)
  
  for (t in 1:tmstep){
    
    Einf = min(rpois(1, (beta * I * S / N)), S) #new infection
    Erecov = min(rpois(1, (I / D)), I) #new recover
    
    S = S - Einf
    I = I + Einf - Erecov
    R = N - S - I
    
    S_rec[t] = S
    I_rec[t] = I
    R_rec[t] = R
    newI_rec[t] = Einf
  
  }
  rec = data.frame(Date=1:tmstep, S=S_rec, I=I_rec, R=R_rec, newI=newI_rec)
  rec
}
```

Run a simulation:

```{r SIR_sim}
N = 1e5 #total population
I = 100 #initial infected population
S = N - I #initial susceptible population
R = 0 #initial recovered population
tmstep = 100 #simulation period (days)
beta = 0.5 #transmission rate
D = 4 #average infectious period

SIRsim = SIR(N, S, I, beta, D, tmstep)

```

Plot results:

```{r plot_SIR_sim, message = FALSE}
library(ggplot2)
Output <- reshape2::melt(SIRsim, id.var='Date')
ggplot(Output, aes(x=Date, y=value, col=variable)) + geom_line()
```

Try different R0:

```{r plot_SIR_sim_R0, message = FALSE}
library(cowplot)
R0=c(0.8,1.0,1.4,2)
pltlst <- list()
for(i in 1:4){
  beta = R0[i]/D
  SIRsim = SIR(N, S, I, beta, D, tmstep)
  Output <- data.frame(Date=SIRsim$Date,newI=SIRsim$newI)
  pltlst[[length(pltlst) + 1]]<-ggplot(Output, aes(x=Date, y=newI)) + geom_line() +
    ggtitle(paste("R0=",toString(R0[i]),sep=""))
}
cowplot::plot_grid(plotlist = pltlst, ncol = 2, labels = "AUTO")
```

## Metapopulation model

A 5-location metapopulation model:

Variables: Total population in 5 locations $N=(10^5,10^5,10^5,10^5,10^5)^T$, susceptible population $S=(9.99\times10^4,10^5,10^5,10^5,10^5)^T$, infected population $I=(100,0,0,0,0)^T$, recovered population $R=(0,0,0,0,0)^T$.

Parameters: The transmission rate $\beta=(0.4,0.4,0.4,0.4,0.4)^T$, average infectious period $D=4$ days.

Human movement: $\mathbf{M}=\{M_{ij}\}_{5\times5}$, 

$M_{ij}$ - human movement from location $j$ to $i$: $i\leftarrow j$. A one-dimensional chain:
\begin{pmatrix}
0 & 1000 & 0 & 0 & 0 \\ 
1000 & 0 & 1000 & 0 & 0 \\ 
0 & 1000 & 0 & 1000 & 0 \\ 
0 & 0 & 1000 & 0 & 1000 \\ 
0 & 0 & 0 & 1000 & 0 
\end{pmatrix}

Equations:

\begin{eqnarray}
  \frac{dS_i}{dt} &=& -\beta_i\frac{S_iI_i}{N_i}+\sum_j\frac{M_{ij}S_j}{N_j}-\sum_j\frac{M_{ji}S_i}{N_i},\\
  \frac{dI_i}{dt} &=& \beta_i\frac{S_iI_i}{N_i}-\frac{I_i}{D}+\sum_j\frac{M_{ij}I_j}{N_j}-\sum_j\frac{M_{ji}I_i}{N_i},\\
  \frac{dR_i}{dt} &=& \frac{I_i}{D}+\sum_j\frac{M_{ij}R_j}{N_j}-\sum_j\frac{M_{ji}R_i}{N_i}.
\end{eqnarray}

Function:

```{r metaSIR}
#metaSIR model
metaSIR<-function(N, S, I, beta, D, M, tmstep){
  #total popualtion: N; susceptible: S; infected: I; 
  #transmission rate: beta; average infectious period: D
  #M: mobility matrix
  #simulation time: tmstep (days)
  
  num_loc = dim(M)[1] #number of locations
  
  #time series for susceptible, infected and newly infected population
  S_rec = I_rec = R_rec = newI_rec = matrix(0,num_loc,tmstep)
  #updated S I and R
  Snew = Inew = Rnew = matrix(0,num_loc,1)
  
  for (t in 1:tmstep){
    
    for (i in 1:num_loc){
      
      Einf = min(rpois(1, (beta[i] * I[i] * S[i] / N[i])), S[i]) #new infection
      Erecov = min(rpois(1, (I[i] / D)), I[i]) #new recover
      
      ESenter = 0 #S entering location i
      ESleft = 0 #S leaving location i
      EIenter = 0 #I entering location i
      EIleft = 0 #I leaving location i
      
      for (j in 1:num_loc){
        ESenter = ESenter + round(M[i,j]*S[j]/N[j])
        ESleft = ESleft + round(M[j,i]*S[i]/N[i])
        EIenter = EIenter + round(M[i,j]*I[j]/N[j])
        EIleft = EIleft + round(M[j,i]*I[i]/N[i])
      }
      
      Snew[i] = S[i] - Einf + ESenter - ESleft
      Inew[i] = I[i] + Einf - Erecov + EIenter -EIleft
      Rnew[i] = N[i] - Snew[i] - Inew[i]
      
      newI_rec[i,t] = Einf
      
    }
    
    S = Snew
    I = Inew
    R = Rnew
    
    S_rec[,t] = Snew
    I_rec[,t] = Inew
    R_rec[,t] = Rnew
  
  }
  rec = list(Location=1:num_loc,Date=1:tmstep,S=S_rec,I=I_rec,R=R_rec,newI=newI_rec)
  rec
}
```

Run a simulation:

```{r metaSIR_sim}
N = matrix(1e5,5,1) #total population
I = matrix(0,5,1) 
I[1] = 100 #initial infected population
S = N - I #initial susceptible population
R = matrix(0,5,1) #initial recovered population
tmstep = 100 #simulation period (days)
num_loc = dim(N)[1]
beta = matrix(0.4,5,1) #transmission rate
D = 4 #average infectious period
M = matrix(c(0,1000,0,0,0,1000,0,1000,0,0,0,1000,0,1000,0,0,0,1000,0,1000,0,0,0,1000,0),nrow=5,ncol=5,byrow=TRUE)

metaSIRsim = metaSIR(N, S, I, beta, D, M, tmstep)

```

Plot results:

```{r plot_metaSIR_sim}
#set column and row names
newI=metaSIRsim[["newI"]]
dimnames(newI) = list(paste0('Location ',seq(1:num_loc)),1:tmstep)
Output <- reshape2::melt(newI)
Output <- data.frame(Output)
colnames(Output) <- c("Location", "Date","value")
ggplot(Output, aes(x=Date, y=value, col=Location)) + geom_line()
```

Change mobility: 10%, 100%, 500% and 1000%

```{r plot_mestSIR_sim_M, message = FALSE}
library(cowplot)
k=c(0.1,1,5,10)
pltlst <- list()
for(i in 1:4){
  Mcnt = k[i]*M
  metaSIRsim = metaSIR(N, S, I, beta, D, Mcnt, tmstep)
  newI=metaSIRsim[["newI"]]
  dimnames(newI) = list(paste0('Location ',seq(1:num_loc)),1:tmstep)
  Output <- reshape2::melt(newI)
  Output <- data.frame(Output)
  colnames(Output) <- c("Location", "Date","value")
  pltlst[[length(pltlst) + 1]]<-ggplot(Output, aes(x=Date, y=value, col=Location)) + geom_line() +
    ggtitle(paste("k=",toString(k[i]),sep=""))
}
cowplot::plot_grid(plotlist = pltlst, ncol = 2, labels = "AUTO")
```


Control in location 2: reduce $\beta$ by 10%, 30%, 50% and 90%

```{r plot_mestSIR_sim_beta, message = FALSE}
library(cowplot)
k=c(0.9,0.7,0.5,0.1)
pltlst <- list()
for(i in 1:4){
  betacnt=beta
  betacnt[2] = betacnt[2]*k[i]
  metaSIRsim = metaSIR(N, S, I, betacnt, D, M, tmstep)
  newI=metaSIRsim[["newI"]]
  dimnames(newI) = list(paste0('Location ',seq(1:num_loc)),1:tmstep)
  Output <- reshape2::melt(newI)
  Output <- data.frame(Output)
  colnames(Output) <- c("Location", "Date","value")
  pltlst[[length(pltlst) + 1]]<-ggplot(Output, aes(x=Date, y=value, col=Location)) + geom_line() +
    ggtitle(paste("beta2*",toString(k[i]),sep=""))
}
cowplot::plot_grid(plotlist = pltlst, ncol = 2, labels = "AUTO")
```
